% ------------------------------------------------------------------
\documentclass[12 pt]{article} % A4 paper set by geometry package below
\pagenumbering{arabic}
\setlength{\parindent}{10 mm}
\setlength{\parskip}{12 pt}

% Nimbus Sans font should be reasonably legible
\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}
\usepackage[T1]{fontenc}  % Without this \textsterling produces $

% Section header spacing
\usepackage{titlesec}
\titlespacing\section{0pt}{12pt plus 4pt minus 2pt}{0pt plus 2pt minus 2pt}
\titlespacing\subsection{0pt}{12pt plus 4pt minus 2pt}{0pt plus 2pt minus 2pt}
\titlespacing\subsubsection{0pt}{12pt plus 4pt minus 2pt}{0pt plus 2pt minus 2pt}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{verbatim}    % For comment
\usepackage[paper=a4paper, marginparwidth=0 cm, marginparsep=0 cm, top=2.5 cm, bottom=2.5 cm, left=3 cm, right=3 cm, includemp]{geometry}
\usepackage[pdftex, pdfstartview={FitH}, pdfnewwindow=true, colorlinks=true, citecolor=blue, filecolor=blue, linkcolor=blue, urlcolor=blue, pdfpagemode=UseNone]{hyperref}

% Put module code and last-modified date in footer
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\cfoot{{\small \thisunit}\hfill \thepage\hfill {\small \moddate}}

% Hopefully address Canvas complaints about pdf tagging
%\usepackage[tagged]{accessibility}
\hypersetup {
  pdfauthor={David Schaich},
  pdftitle={Statistical Physics Tutorial},
}
% ------------------------------------------------------------------



% ------------------------------------------------------------------
% Shortcuts
\newcommand{\Rbb}{\ensuremath{\mathbb R} }
\newcommand{\De}{\ensuremath{\Delta} }
% ------------------------------------------------------------------



% ------------------------------------------------------------------
\begin{document}
\newcommand{\thisunit}{MATH327 Tutorial (CLT)}
\newcommand{\moddate}{Last modified 30 Jan.~2023}
\begin{center}
  {\Large \textbf{MATH327: Statistical Physics, Spring 2023}} \\[12 pt]
  {\Large \textbf{Tutorial activity \ --- \ Central limit theorem}} \\[24 pt]
\end{center}

To illustrate the central limit theorem, let's consider the roulette wheel introduced in class (and on page 13 of the lecture notes).
A simple game of roulette would let us place bets on whether or not the ball will end up in a red- or black-coloured pocket: If we bet correctly we get back twice the money we put in; otherwise we lose our money.
We'll define our (potentially negative) \textit{gain} to be the amount we receive minus the amount we spend on bets.

\begin{itemize}
  \item Suppose we place \textsterling5 bets on `black' for each of $N$ spins of the roulette wheel.
        What are the probabilities and gains of winning and of losing for any single one of those spins?
        Letting $W = 0, \cdots, N$ be the number of spins where we win, what is our total gain $G_W$ as a function of $(N, W)$?
  \item Recall that the number of different ways we could win $W$ times out of $N$ attempts is given by the binomial coefficient
        \begin{equation*}
          \binom{N}{W} = \frac{N!}{W! \; (N - W)!},
        \end{equation*}
        with $0! = 1$.
        Setting $N = 5$, what are the six probabilities $p_0$ through $p_5$ that we win $W = 0, \cdots, 5$ times?
        What is the general $p_W$ for any $(N, W)$?
  \item Now let's apply the central limit theorem to this setup.
        What are the mean gain and its variance for a single spin of the wheel?
        What is the resulting probability distribution $p(g)$ given by the central limit theorem for the gain $g \in \Rbb$ after $N \gg 1$ spins?
  \item In order to compare the distribution $p(g)$ against the probabilities $p_W$ considered above, we need to integrate $p(g)$ over appropriate intervals as discussed in class (and on page 16 of the lecture notes).
        Natural intervals to consider are
        \begin{equation*}
          P_{\text{integ}}(G_W) \equiv \int_{G_W - \De G / 2}^{G_W + \De G / 2} p(g) \; dg,
        \end{equation*}
        where $\De G = G_{W + 1} - G_W$ is a constant you can read off from your work so far.
        These numerical integrations are not convenient to do by hand, but can easily be performed by Maple, Python, MATLAB, Mathematica, etc.
        Alternately, we can simplify further by approximating $p(g)$ as a constant within each interval, which would give us
        \begin{equation*}
          P_{\text{const}}(G_W) \equiv p(G_W) \; \De G.
        \end{equation*}
        Keeping $N = 5$, what are the six $P_{\text{integ}}$ and $P_{\text{const}}$?
\end{itemize}

\end{document}
% ------------------------------------------------------------------
